---
title: "Reproducible Research - Peer Assessment 2"
author: "Kevin E. D'Elia"
date: "09/18/2015"
output: 
  html_document:
    keep_md: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of NOAA Storm Database
Access, data-wrangle, and provide data visualization of key variables contained in the NOAA Storm Database.


# Synopsis of Analysis
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. [NEED MORE WORDS HERE!!!]

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.
Data


# Data Processing

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from this link:
[Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

[National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

[National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

Once the data file has been downloaded, it can be loaded into **R** using the following function call:

```{r load_data, cache=TRUE}
stormData <- read.csv(bzfile("./data/repdata_data_StormData.csv.bz2"), stringsAsFactors = FALSE)
```

A quick display of the dataframe's variables will help in subsequent selection and filtering approaches:

```{r descriptive_statistics}
str(stormData)
```

There are 37 variables in this dataset, and only a small subset of those are needed for this analysis.  So, load the `dplyr` package in order to use its elegant selection and filtering capabilities.

```{r load_dplyr}
suppressMessages(library(dplyr))
library("dplyr")
```

The goal of this assignment is to answer the following questions:

1. Across the United States, which types of events (as indicated in the `EVTYPE` variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

So, for the public health question, the **FATALITIES** and **INJURIES** columns would be of significance and for the economic question, the **PROPDMG**, **PROPDMGEXP**, **CROPDMG**, and **CROPDMGEXP** columns will contain useful information.  Naturally, the **EVTYPE** will be included, since it is the categorical value of primary concern.

```{r select_columns}
stormData  <- stormData  %>% select(c(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
```

There are `r nrow(stormData)` rows in the original dataset and, again, not all of those are needed, since some cells may contain values not useful in the analysis.  Furthermore, we can massage the data into dataframes that are more focused on the questions we'd like to answer, so we can select from our first subset only those columns relevant to each of our two questions and put them in separate dataframes.  Finally, for the population health question, since we are interested in events that caused fatalities and injuries only, let's filter out ones that didn't cause any:

```{r filter_data_for_population}
populationData <-   stormData  %>% 
                    select(c(EVTYPE, FATALITIES, INJURIES)) %>% 
                    filter(FATALITIES > 0.00 & INJURIES > 0.00)
```

Repeat the process, this time for the economic question.  We need only those columns related to property and crop damage; the exponent column is a multiplier which will come into play later on in the analysis.  For events that didn't cause any property or crop damage, exclude them:

```{r filter_data_for_economics}
economicData <-   stormData  %>% 
                  select(c(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)) %>% 
                  filter(PROPDMG > 0.00 & CROPDMG > 0.00)
```

This leaves `r nrow(populationData)` rows for the dataset used to answer the population question and `r nrow(economicData)` rows to answer the economic question, a significant reduction from the `r nrow(stormData)` rows of the original dataset.  But, can the datasets be reduced even further?  If we run `summary` against columns of interest in each dataset, we may get some guidance.

```{r run_summary_statistics_for_population}
summary(populationData$FATALITIES)
summary(populationData$INJURIES)
```

Looking at these results, the significant values appear in the 3rd quartile.  Using that information, let's filter the population dataset down a bit more:

```{r filter_population_data_based_on_summary}
populationData  <- populationData  %>% filter(FATALITIES >= 2 & INJURIES >= 20)
```

Appying the same approach to the economic data, we have the following summary information:

```{r run_summary_statistics_for_economic}
summary(economicData$PROPDMG)
summary(economicData$CROPDMG)
```

Again, it appears that the dataset can be further reduced by using the values for the 3rd quartile, and so we apply that filter now:

```{r filter_economic_data_based_on_summary}
economicData  <- economicData  %>% filter(PROPDMG >= 50 & CROPDMG >= 50)
```

The next step in the processing is to reverse-map the 48 Event Types defined by the National Weather Service (NWS) to the entries in our datasets.  The complete list can be found in the link to the National Weather Service Storm Data Documentation given above and is reproduced here for completeness.  

Event Type            | Event Type             | Event Type              | Event Type
----------------------|------------------------|-------------------------|-----------
Astronomical Low Tide |Extreme Cold/Wind Chill |Hurricane (Typhoon)      |Storm Surge/Tide
Avalanche             |Flash Flood             |Ice Storm                |Strong Wind
Blizzard              |Flood                   |Lake-Effect Snow         |Thunderstorm Wind
Coastal Flood         |Frost/Freeze            |Lakeshore Flood          |Tornado
Cold/Wind Chill       |Funnel Cloud            |Lightning                |Tropical Depression
Debris Flow           |Freezing Fog            |Marine Hail              |Tropical Storm
Dense Fog             |Hail                    |Marine High Wind         |Tsunami
Dense Smoke           |Heat                    |Marine Strong Wind       |Volcanic Ash
Drought               |Heavy Rain              |Marine Thunderstorm Wind |Waterspout 
Dust Devil            |Heavy Snow              |Rip Current              |Wildfire
Dust Storm            |High Surf               |Seiche                   |Winter Storm
Excessive Heat        |High Wind               |Sleet                    |Winter Weather

Some of the existing `EVTYPE` values are 1::1 with the NWS types but for those which are not, the following process will normalize them.  First, for the population data:

**Note**: This [link](http://forecast.weather.gov/glossary.php?word=tstm) will give the definition for **TSTM**, i.e., thunderstorm, as justification for changing any `EVTYPE` entries with that value to one of Thunderstorm.

```{r map_population_data}
cold  <- grepl("cold", populationData$EVTYPE, ignore.case = T)
populationData$EVTYPE[cold]  <- "Cold/Wind Chill"

rain  <- grepl("rain", populationData$EVTYPE, ignore.case = T)
populationData$EVTYPE[rain]  <- "Heavy Rain"

fog  <- grepl("fog", populationData$EVTYPE, ignore.case = T)
populationData$EVTYPE[fog]  <- "Dense Fog"

heat  <- grepl("heat wave", populationData$EVTYPE, ignore.case = T)
populationData$EVTYPE[heat]  <- "Excessive Heat"

wind  <- grepl("high wind", populationData$EVTYPE, ignore.case = T)
populationData$EVTYPE[wind]  <- "High Wind"

tropical  <- grepl("tropical storm", populationData$EVTYPE, ignore.case = T)
populationData$EVTYPE[tropical]  <- "Tropical Storm"

thunder  <- grepl("tstm", populationData$EVTYPE, ignore.case = T)
populationData$EVTYPE[thunder]  <- "THUNDERSTORM WIND"

stream  <- grepl("urban", populationData$EVTYPE, ignore.case = T)
populationData$EVTYPE[stream]  <- "Flash Flood"

fire  <- grepl("fire", populationData$EVTYPE, ignore.case = T)
populationData$EVTYPE[fire]  <- "Wildfire"

hurricane  <- grepl("hurricane", populationData$EVTYPE, ignore.case = T)
populationData$EVTYPE[hurricane]  <- "Hurricane (Typhoon)"

spout  <- grepl("waterspout", populationData$EVTYPE, ignore.case = T)
populationData$EVTYPE[spout]  <- "Waterspout"
```

Next, for the economic data:

```{r map_economic_data}
dust  <- grepl("dust", economicData$EVTYPE, ignore.case = T)
economicData$EVTYPE[dust] <- "Dust Storm"

cold <- grepl("extreme cold", economicData$EVTYPE, ignore.case = T)
economicData$EVTYPE[cold] <- "Extreme Cold/Wind Chill"

flash <- grepl("flash", economicData$EVTYPE, ignore.case = T)
economicData$EVTYPE[flash] <- "Flash Flood"

FLOOD <- grepl("FL.*D", economicData$EVTYPE, ignore.case = F)
economicData$EVTYPE[FLOOD] <- "Flood"

freeze <- grepl("freeze", economicData$EVTYPE, ignore.case = T)
economicData$EVTYPE[freeze] <- "Frost/Freeze"

gust <- grepl("gusty", economicData$EVTYPE, ignore.case = T)
economicData$EVTYPE[gust] <- "High Wind"

hail <- grepl("hail", economicData$EVTYPE, ignore.case = T)
economicData$EVTYPE[hail] <- "Hail"

heat <- grepl("heat", economicData$EVTYPE, ignore.case = T)
economicData$EVTYPE[heat] <- "Heat"

storm <- grepl("typhoon", economicData$EVTYPE, ignore.case = T)
economicData$EVTYPE[storm] <- "Tropical Storm"

rain <- grepl("rain", economicData$EVTYPE, ignore.case = T)
economicData$EVTYPE[rain] <- "Heavy Rain"

hurricane <- grepl("hurricane", economicData$EVTYPE, ignore.case = T)
economicData$EVTYPE[hurricane] <- "Hurricane (Typhoon)"

thunder <- grepl("thu.*s.*", economicData$EVTYPE, ignore.case = T)
economicData$EVTYPE[thunder] <- "Thunderstorm Wind"

tstm <- grepl("tstm", economicData$EVTYPE, ignore.case = T)
economicData$EVTYPE[tstm] <- "Thunderstorm Wind"

tropical <- grepl("tropical", economicData$EVTYPE, ignore.case = T)
economicData$EVTYPE[tropical] <- "Tropical Storm"

wildfire <- grepl("wildfire", economicData$EVTYPE, ignore.case = T)
economicData$EVTYPE[wildfire] <- "Wildfire"

forest <- grepl("forest", economicData$EVTYPE, ignore.case = T)
economicData$EVTYPE[forest] <- "Wildfire"

winter  <- grepl("winter", economicData$EVTYPE, ignore.case = T)
economicData$EVTYPE[winter] <- "Winter Storm"
```

A little more cleanup:

1. make the `EVTYPE` values all free of extraneous whitespace and in upper case
2. make `EVTYPE` a factor variable now so its values can be used in later graphs

```{r clean_factorize}
populationData$EVTYPE  <- toupper(trimws(populationData$EVTYPE))
populationData$EVTYPE  <- as.factor(populationData$EVTYPE)

economicData$EVTYPE <- toupper(trimws(economicData$EVTYPE))
economicData$EVTYPE  <- as.factor(economicData$EVTYPE)
```

As the final step in the cleanup, we need to examine the multiplier column for economic data.  The `PROPDMGEXP` and `CROPDMGEXP` can have a wide range of values ([see here](https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html) for more information), many of which may have been already culled out as a result of the data-wrangling process.  Let's see what's left:

```{r examine_exponents}
unique(economicData$PROPDMGEXP)
unique(economicData$CROPDMGEXP)
```

So for the `PROPDMGEXP` we have just multipliers for thousands and millions of dollars; the `CROPDMGEXP` needs a bit more cleanup.  First, we'll remove the **0** multiplier, since it is not a contributor to the overall damage costs, then we'll normalize the **k** value to be consistent with the other `CROPDMGEXP` values.

```{r adjust_multipliers}
economicData <- economicData %>% filter(CROPDMGEXP != 0)
economicData$CROPDMGEXP <- toupper(economicData$CROPDMGEXP)
```

Now we can start summarizing the data and arriving at some results.

# Results

Once the data has been processed into a format suitable for analysis, we can summarize the totals for fatalities and injuries based on each distinct `EVTYPE`.

```{r summarize_population_data}
populationTotals <- populationData %>% 
                    group_by(EVTYPE) %>% 
                    summarise(fatalities = sum(FATALITIES), injuries = sum(INJURIES))
```

Now show the plot for this data:

```{r plot_population, fig.path="./figures/", fig.cap="Figure 1: Summary of fatalities and injuries by Event Type", results="asis", fig.width=12}
plot(populationTotals$fatalities, 
     type="l", 
     col="blue", 
     axes=F, 
     ann=T, 
     xlab="", 
     ylab="Totals", 
     lwd=2,
     ylim=c(0, 5000))
axis(1, 
     at=1:20, 
     labels = F)
axis(2, 
     las=1, 
     cex.axis=0.8)
text(1:20, 
     par("usr")[3] - 2, 
     srt=45, 
     adj=1, 
     labels=populationTotals$EVTYPE, 
     xpd=T, 
     cex=0.7)
lines(populationTotals$injuries, 
      col="orange", 
      lwd=2)
points(populationTotals$fatalities, pch=18, cex=0.8)
points(populationTotals$injuries, pch=20, cex=0.8)
legend("topright", 
      c("fatalities","injuries"), 
      cex=0.8, 
      col=c("blue","orange"), 
      lty=1,
      pch=c(18, 20))
box()
```

******

From the table and the graphic, it is clear that Tornadoes are the primary cause of fatalities and injuries, followed by Excessive Heat, respectively.

Now we examine the question of which event cost the most in terms of property and crop damage.  To do that, we use the multiplier associated with each indicator.  For property, we do this:

```{r calculate_property_damage_costs}
economicData$PROPDMGCOST <- 
  ifelse(economicData$PROPDMGEXP == "K", 
         economicData$PROPDMG * 1.0e+03, 
         economicData$PROPDMG * 1.0e+06)
```

For crop damage, we calculate it this way:

```{r calculate_crop_damage_costs}
economicData$CROPDMGCOST <- 
  ifelse(economicData$CROPDMGEXP == "K", 
         economicData$CROPDMG * 1.0e+03, 
         economicData$CROPDMG * 1.0e+06)
```

We need only the `EVTYPE` and calculated columns now, so select only those:
```{r clean_up}
economicData <- economicData %>% 
                select(c(EVTYPE, PROPDMGCOST, CROPDMGCOST))
```

```{r summarize_economic_data}
economicCosts <-  economicData %>% 
                  group_by(EVTYPE) %>% 
                  summarise(propertyCosts = sum(PROPDMGCOST), cropCosts = sum(CROPDMGCOST))
```

```{r plot_property_costs, fig.path="./figures/", fig.cap="Figure 2: Property Damage Costs by Event Type", results="asis", fig.width=12}
plot(economicCosts$propertyCosts, 
     type="l", 
     col="blue", 
     axes=F, 
     ann=T, 
     xlab="", 
     ylab="Total Cost Property Damage (US Dollars)", 
     cex.lab=0.6, 
     lwd=2)
axis(1, 
     at=1:nrow(economicCosts), 
     labels = F)
axis(2, 
     las=1, 
     cex.axis=0.8)
text(1:nrow(economicCosts), 
     par("usr")[3] - 2, 
     srt=45, 
     adj=1, 
     labels=economicCosts$EVTYPE, 
     xpd=T, 
     cex=0.7)
points(economicCosts$propertyCosts, pch=18)
box()
```

```{r plot_crop_costs, fig.path="./figures/", fig.cap="Figure 3: Crop Damage Costs by Event Type", results="asis", fig.width=12}
plot(economicCosts$cropCosts, 
     type="l", 
     col="orange", 
     axes=F, 
     ann=T, 
     xlab="", 
     ylab="Total Cost Crop Damage (US Dollars)", 
     cex.lab=0.6, 
     lwd=2)
axis(1, 
     at=1:nrow(economicCosts), 
     labels = F)
axis(2, 
     las=1, 
     cex.axis=0.8)
text(1:nrow(economicCosts), 
     par("usr")[3] - 2, 
     srt=45, 
     adj=1, 
     labels=economicCosts$EVTYPE, 
     xpd=T, 
     cex=0.7)
points(economicCosts$cropCosts, pch=18)
box()
```